---
title: "Build an App"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build an App}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

While `querychat_app()` provides a quick way to start exploring data, building bespoke Shiny apps with QueryChat unlocks the full power of integrating natural language data exploration with custom visualizations, layouts, and interactivity. This guide shows you how to integrate QueryChat into your own Shiny applications and leverage its reactive data outputs to create rich, interactive dashboards.

## Starter template

Integrating QueryChat into a Shiny app requires just three steps:

1. Initialize a `QueryChat` instance with your data
2. Add the QueryChat UI component (either `$sidebar()` or `$ui()`)
3. Use reactive values like `$df()`, `$sql()`, and `$title()` to build outputs that respond to user queries

Here's a starter template demonstrating these steps:

```{r}
library(shiny)
library(bslib)
library(querychat)
library(DT)
library(palmerpenguins)

# Step 1: Initialize QueryChat
qc <- QueryChat$new(penguins)

# Step 2: Add UI component
ui <- page_sidebar(
  sidebar = qc$sidebar(),
  card(
    card_header("Data Table"),
    dataTableOutput("table")
  ),
  card(
    fill = FALSE,
    card_header("SQL Query"),
    verbatimTextOutput("sql")
  )
)

# Step 3: Use reactive values in server
server <- function(input, output, session) {
  qc_vals <- qc$server()
  
  output$table <- renderDataTable({
    datatable(qc_vals$df(), fillContainer = TRUE)
  })

  output$sql <- renderText({
    qc_vals$sql() %||% "SELECT * FROM penguins"
  })
}

shinyApp(ui, server)
```

::: {.alert .alert-info}
You'll need to call the `qc$server()` method within your server function to set up QueryChat's reactive behavior, and capture its return value to access reactive data.
:::

## Reactives

There are three main reactive values provided by QueryChat for use in your app:

### Filtered data {#filtered-data}

The `$df()` method returns the current filtered and/or sorted data frame. This updates whenever the user prompts a filtering or sorting operation through the chat interface (see [Data updating](tools.html#data-updating) for details).

```{r}
qc_vals <- qc$server()

output$table <- renderDataTable({
  qc_vals$df()  # Returns filtered/sorted data
})
```

You can use `$df()` to power any output in your app - visualizations, summary statistics, data tables, and more. When a user asks to "show only Adelie penguins" or "sort by body mass", `$df()` automatically updates, and any outputs that depend on it will re-render.

### SQL query {#sql-query}

The `$sql()` method returns the current SQL query as a string. This is useful for displaying the query to users for transparency and reproducibility:

```{r}
qc_vals <- qc$server()

output$current_query <- renderText({
  qc_vals$sql() %||% "SELECT * FROM penguins"
})
```

You can also use `$sql()` as a setter to programmatically update the query (see [Programmatic filtering](#programmatic-filtering) below).

### Title {#title}

The `$title()` method returns a short description of the current filter, provided by the LLM when it generates a query. For example, if a user asks to "show Adelie penguins", the title might be "Adelie penguins".

```{r}
qc_vals <- qc$server()

output$card_title <- renderText({
  qc_vals$title() %||% "All Data"
})
```

Returns `NULL` when no filter is active. You can also use `$title()` as a setter to update the title programmatically.

## Custom UI

In the starter template above, we used the `$sidebar()` method for a simple sidebar layout. In some cases, you might want to place the chat UI somewhere else in your app layout, or just more fully customize what goes in the sidebar. The `$ui()` method is designed for this -- it returns the chat component without additional layout wrappers.

For example, you might want to create some additional controls to [reset filters](#programmatic-filtering) alongside the chat UI:

```{r}
library(querychat)
library(palmerpenguins)

qc <- QueryChat$new(penguins)

ui <- page_sidebar(
  sidebar = sidebar(
    qc$ui(),  # Chat component
    actionButton("reset", "Reset Filters", class = "w-100"),
    fillable = TRUE,
    width = 300
  ),
  # Main content here
)
```

::: {.alert .alert-info}
**Customizing chat UIs**

See `{shinychat}`'s [docs](https://posit-dev.github.io/shinychat/r/index.html) to learn more about customizing the chat UI component returned by `qc$ui()`.
:::

## Data views

Thanks to Shiny's support for interactive visualizations with packages like [plotly](https://plotly.com/r/), it's straightforward to create rich data views that depend on QueryChat data. Here's an example of an app showing both the filtered data and a bar chart depending on that same data:

```{r}
library(shiny)
library(bslib)
library(querychat)
library(DT)
library(plotly)
library(dplyr)
library(palmerpenguins)

qc <- QueryChat$new(penguins)

ui <- page_sidebar(
  sidebar = qc$sidebar(),
  layout_columns(
    card(
      card_header("Data Table"),
      dataTableOutput("table")
    ),
    card(
      card_header("Body Mass by Species"),
      plotlyOutput("mass_plot")
    )
  )
)

server <- function(input, output, session) {
  qc_vals <- qc$server()

  output$table <- renderDataTable({
    qc_vals$df()
  })

  output$mass_plot <- renderPlotly({
    d <- qc_vals$df() %>%
      group_by(species) %>%
      summarise(avg_mass = mean(body_mass_g, na.rm = TRUE))

    ggplot(d, aes(x = species, y = avg_mass, fill = species)) +
      geom_col() +
      labs(x = "Species", y = "Average Body Mass (g)") +
      theme_minimal()
  })
}

shinyApp(ui, server)
```

Now when a user filters the data through natural language (e.g., "filter to only male penguins"), both the table and the chart update automatically.

## Programmatic updates {#programmatic-filtering}

QueryChat's reactive state can be updated programmatically. For example, you might want to add a "Reset Filters" button that clears any active filters and returns the data table to its original state. You can do this by setting both the SQL query and title to their default values. This way you don't have to rely on both the user and LLM to send the right prompt.

```{r}
ui <- page_sidebar(
  sidebar = sidebar(
    qc$ui(),
    hr(),
    actionButton("reset", "Reset Filters")
  ),
  # Main content
  card(dataTableOutput("table"))
)

server <- function(input, output, session) {
  qc_vals <- qc$server()

  output$table <- renderDataTable({
    qc_vals$df()
  })

  observeEvent(input$reset, {
    qc_vals$sql("")
    qc_vals$title(NULL)
  })
}

shinyApp(ui, server)
```

This is equivalent to the user asking the LLM to "reset" or "show all data".

## Multiple datasets

You can use multiple QueryChat instances in a single app to explore different datasets. Just ensure each instance has a different table name (or `id` which derives the table name) to avoid conflicts. Here's an example with two datasets:

```{r}
library(shiny)
library(bslib)
library(querychat)

penguins_data <- palmerpenguins::penguins
mtcars_data <- mtcars

qc_penguins <- QueryChat$new(penguins_data, table_name = "penguins")
qc_mtcars <- QueryChat$new(mtcars_data, table_name = "mtcars")

ui <- page_navbar(
  title = "Multiple Datasets",
  sidebar = sidebar(
    id = "sidebar",
    conditionalPanel(
      "input.navbar == 'Penguins'",
      qc_penguins$ui()
    ),
    conditionalPanel(
      "input.navbar == 'Cars'",
      qc_mtcars$ui()
    )
  ),
  nav_panel(
    "Penguins",
    card(dataTableOutput("penguins_table"))
  ),
  nav_panel(
    "Cars",
    card(dataTableOutput("mtcars_table"))
  ),
  id = "navbar"
)

server <- function(input, output, session) {
  qc_penguins_vals <- qc_penguins$server()
  qc_mtcars_vals <- qc_mtcars$server()

  output$penguins_table <- renderDataTable({
    qc_penguins_vals$df()
  })

  output$mtcars_table <- renderDataTable({
    qc_mtcars_vals$df()
  })
}

shinyApp(ui, server)
```

## See also

- [Greet users](greet.html) - Create welcoming onboarding experiences
- [Provide context](context.html) - Help the LLM understand your data better
- [Tools](tools.html) - Understand what QueryChat can do under the hood
